Adaptive Sequential Model Selection

### 2015-07-01 paper.tex
rob

\newcommand*\mystrut{\vrule width0pt height0pt depth1.5ex\relax}
\newcommand{\underlabel}{\underbracket[1pt][.5pt]{\mystrut \quad\;\; \sub \quad\;\; }}
\newcommand{\JTcomment}[1]{{\color{blue}{(JT: \bf \sc #1) }}}
\newcommand{\WFcomment}[1]{{\color{red}{(WF: \bf \sc #1) }}}

@@

\section{Introduction}

Many model selection procedures produce  a sequence of increasingly complex models. Often our goal is to choose the simplest model that is not falsified by the available data.
In other words, we seek  the index of the smallest adequate model ---  the smallest model that cannot be falsified given available evidence.
Examples of such model selection procedures include  forward stepwise linear regression, least angle regression (LAR) and  the ever-active path in $\ell_!$-regularized models .
All of these were studied in \citet{taylor2014exact}. Another, different, example is principal components analysis in which we try to determine the correct rank of the
covariance matrix. This problem was studied in \citet{choi2014selecting}.

@@

   \hline\\
\end{tabular}
\caption[tab:diab]{\em Forward stepwise regression for the diabetes data: naive p-values from forward stepwise regression, 
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper. The red annotation indicates the stopping step
prescribed by {\tt forwardStop}. For the max-t p-values, this gives exact FDR control at the 10\% level.}
\label{tab:diab}
\end{table}
The table shows the naive p-values from forward stepwise regression, that do not account for the selection, as well as the
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper.
The latter two sets of p-values test someone different hypotheses:  the p-values from the saturated model test whether the partial regression of the predictor just added is zero,
what we call the ``incremental null''
In contrast, the max-t pvalues test whether the current model is adequate, that is, all predictors not in the model  have coefficients equal to zero,
the ``complete null''. (The issue of what hypothesis to test is discussed in section \ref{sec:whichnull}.)
Now it turns out that the max-t test if often more powerful that the saturated model test, and importantly, the max-t pvalues are exactly
independent after we reach the point the model sequence when the complete null is true. This independence means that we can apply
sequential FDR exact stopping rules, such as ``forwardStop'' (\citet{sell2013sequential}). When applied to the max-t pvalues at FDR 0.10, {\tt forwardStop}
suggested stopping at step 8. In constant, the saturated lvalues gave stopping a point of 3, with no guarantee of FDR control
due to the lack of independence of the p-values.

An outline of this paper is as follows:

OUTLINE

@@

\subsection{Notation and Problem Setting}\label{sec:genericSetting}

@@

\section{Sequential Inference}\label{sec:sequential}

@@

\citet{gsell2013sequential} propose another stopping rule that controls the FWER, which they called {\em strong stop}. Define
\[
  \hk_{S}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    \exp\left(\sum_{i=k}^d \frac{\log p_i}{i}\right) 
    \leq \frac{\alpha k}{d}\right\}
\]
Even if $p_k$ is a little larger than $\alpha$, say $\alpha=0.05$ and $p_k=0.06$, StrongStop can still reject $M_{k-1}$ if the next three or four $p$-values are also relatively small.

@@

\citet{gsell2013sequential} propose another stopping rule, {\em forward stop}, that controls the FDR:
\[
  \hk_{F}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    -\frac{1}{k}\sum_{i=1}^k \log(1-p_i) \leq \alpha\right\}
\]
If $p_k$ is uniform then  $-\log(1-p_k)$ is an exponential random variable. If all of the null $p$-values are uniform and the others are zero, then only the null ones contribute to the sum, and
\[
\widehat{\text{FDR}}_k = -\frac{1}{k}\sum_{i=1}^k \log(1-p_i)
\]
is a Gamma random variable with mean $V/k$ and variance $V/k^2$. We stop the last time the estimate of FDR is less than $\alpha$.

@@

\WFcomment{Discuss Rina's stuff.}

@@

\section{Inference for One Step}\label{sec:singleStep}

@@

\section{Computation}\label{sec:computation}

@@

 Let $F_{\mu,\sigma^2}^{a,b}(t)$ be the  cumulative distribution function of the truncated normal.
 Finally let $T=|T_{j^*}|$.
Then by assuming independence, a  simple approximation to the max-t value is given by
\begin{equation}
1-\prod_{j\notin E} \{F_{0,||\eta||^2}^{v_j^-, v_j^+}(T)-   F_{0,||\eta||^2}^{v_j^-, v_j^+}(-T) \}
\label{eqn:approxpv}
\end{equation}
Figure \ref{fig:approxpv} shows an example with $n=50$ and $p=10$ features having pairwise correlation 0.5.
There are two moderate signals.
Shown are the p-values from the ``exact'' numerical computation (with $100,000$ bootstrap samples)
and the approximation based on (\ref{eqn:approxpv}).
\begin{figure}[htp]
\centering
  \includegraphics[width=3in]{figs/approxpv.pdf}
  \caption{Exact max-t pvalues and approximation based on (\ref{eqn:approxpv}), for a simulated example
  with $n=50, p=10$.}
  \label{fig:approxpv}
\end{figure}
The approximation seems quite good. By Sidak's lemma, we might conjecture that the approximation is conservative, but we have no proof of this.

\section{Simulation: Sparse Linear Regression}\label{sec:sparseReg}

@@

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Fri May  1 09:29:28 2015
\begin{table}[ht]
  \centering
  \begin{tabular}{llcccc}
    \hline
    Method & Stopping Rule & $p_{\text{screen}}$ & $\text{FWER}_{\text{mod}}$ 
    & $\text{FDR}_{\text{mod}}$ 
    & $\text{FDR}_{\text{var}}$ \\ 
    \hline
    Selected & Basic & .290 & .000 & .002 & .039 \\ 
    Selected & Forward & .559 & .027 & .020 & .066 \\ 
    Selected & Strong & .041 & .014 & .008 & .042 \\ 
    \hline
    Saturated & Basic & .000 & .000 & .000 & .028 \\ 
    Saturated & Forward & .014 & .000 & .000 & .030 \\ 
    Saturated & Strong & .000 & .000 & .000 & .032 \\ 
    \hline
    Knockoffs & & .000 & --- & --- & .231 \\ 
    \hline
  \end{tabular}
  \caption{\WFcomment{Comment on results once they are correct. These numbers are wrong right now.}}
\end{table}

\section{Simulation: Principal Components Analysis}\label{sec:pca}

\WFcomment{Can we do this one?}

\section{Discussion}

>>>>

\newcommand*\mystrut{\vrule width0pt height0pt depth1.5ex\relax}
\newcommand{\underlabel}{\underbracket[1pt][.5pt]{\mystrut \quad\;\; \sub \quad\;\; }}
\newcommand{\JTcomment}[1]{{\color{blue}{(JT: \bf \sc #1) }}}
\newcommand{\WFcomment}[1]{{\color{red}{(WF: \bf \sc #1) }}}
\newcommand{\RTcomment}[1]{{\color{red}{(RT: \bf \sc #1) }}}

@@

\section{Introduction}
\label{sec:introduction}
Many model selection procedures produce  a sequence of increasingly complex models. Often our goal is to choose the simplest model that is not falsified by the available data.
In other words, we seek  the index of the smallest adequate model ---  the smallest model that cannot be falsified given available evidence.
Examples of such model selection procedures include  forward stepwise linear regression, least angle regression (LAR) and  the ever-active path in $\ell_!$-regularized models .
All of these were studied in \citet{taylor2014exact}. Another, different, example is principal components analysis in which we try to determine the correct rank of the
covariance matrix. This problem was studied in \citet{choi2014selecting}.

@@

   \hline\\
\end{tabular}
\caption[tab:diab]{\em Forward stepwise regression for the diabetes data: naive p-values from forward stepwise regression, 
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper. The red annotation indicates the stopping step
prescribed by {\tt forwardStop}. For the max-t p-values, this gives exact FDR control at the 10\% level.}
\label{tab:diab}
\end{table}
The table shows the naive p-values from forward stepwise regression, that do not account for the selection, as well as the
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper.
The latter two sets of p-values test someone different hypotheses:  the p-values from the saturated model test whether the partial regression of the predictor just added is zero,
what we call the ``incremental null''
In contrast, the max-t pvalues test whether the current model is adequate, that is, all predictors not in the model  have coefficients equal to zero,
the ``complete null''. (The issue of what hypothesis to test is discussed in section \ref{sec:whichnull}.)
Now it turns out that the max-t test if often more powerful that the saturated model test, and importantly, the max-t pvalues are exactly
independent after we reach the point the model sequence when the complete null is true. This independence means that we can apply
sequential FDR exact stopping rules, such as ``forwardStop'' (\citet{gsell2013sequential}). When applied to the max-t pvalues at FDR 0.10, {\tt forwardStop}
suggested stopping at step 8. In constant, the saturated lvalues gave stopping a point of 3, with no guarantee of FDR control
due to the lack of independence of the p-values.

An outline of this paper is as follows:
In Section \ref{sec:genericSetting} we setup the problem and give some important definitions. In Section \ref{sec:sequential} we review recently proposed methods for
testing ordered hypothesis using from a sequence of p-values. These method require that the p-values be uniform and independent under the null, and we derive general conditions
to ensure this property. In Section \ref{sec:single step}  we  show how to construct valid selective $p$-values for a single step in the path and derive the max-t test
for  forward stepwise regression and the Next-$\lambda$ test for the lasso. Section \ref{sec:computation}  describes our computational strategy for estimating the p-values using accept-reject and hit and run sampling. We also relate the method to parametric bootstrap sampling. We provide a simulation study in Section \ref{sec:simulation} and briefly discuss applications of our framework to
principal components analysis in section\ref{sec:pac}. The paper ends with a Discussion.

@@

\section{Notation and Problem Setting}
\label{sec:genericSetting}

@@

\section{Sequential Inference}
\label{sec:sequential}

@@

\citet{gsell2013sequential} propose a stopping rule that controls the FWER, which they called {\em strong stop}. Define
\[
  \hk_{S}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    \exp\left(\sum_{i=k}^d \frac{\log p_i}{i}\right) 
    \leq \frac{\alpha k}{d}\right\}
\]
StrongStop uses all of the p-values after step $k$ in its decision to reject at step $k$. 
Even if $p_k$ is a little larger than $\alpha$, say $\alpha=0.05$ and $p_k=0.06$, StrongStop can still reject $M_{k-1}$ if the next three or four $p$-values are also relatively small.

@@

\citet{gsell2013sequential} propose another stopping rule, {\em forward stop}, that controls the FDR:
\[
  \hk_{F}(Y) = \max\left\{k \in \{1,\ldots,d\} :\;
    -\frac{1}{k}\sum_{i=1}^k \log(1-p_i) \leq \alpha\right\}
\]
ForwardStop uses the p-values{\it before} step $k$ in its decision to reject at step $k$. 
If $p_k$ is uniform then  $-\log(1-p_k)$ is an exponential random variable. If all of the null $p$-values are uniform and the others are zero, then only the null ones contribute to the sum, and
\[
\widehat{\text{FDR}}_k = -\frac{1}{k}\sum_{i=1}^k \log(1-p_i)
\]
is a Gamma random variable with mean $V/k$ and variance $V/k^2$. We stop the last time the estimate of FDR is less than $\alpha$.

@@

\citet{li2015accumulation} generalize the ForwardStop rule, introducing the family of {\em accumulation tests}, defined as follows.
Fix a function $h: [0,1] \rightarrow [0,\infty]$ satisfying $\int _{t=0}^1 h(t)dt=1$, called an ``accumulation function''. They define an estimate of the
false discovery proportion at a cutoff $k$ as
\begin{equation}
\widehat{\rm FDP}_h(k)= \frac{\sum_{i=1}^k h(p_i)}{k}
\end{equation}
and then define the adaptive cutoff,  for FDR level $\alpha$, to be                                                                                                                 as
\begin{equation}
\hat k_h={\rm max} \Bigl\{k\in \{1,2,\ldots n\}; \widehat{\rm FDP}_h(k) \leq \alpha \Bigr\}
\end{equation}
The ForwardStop rule falls into this class:  another  choice they study is the ``HingeExp'' function
\begin{equation}
h_{\rm HingeExp}(t)= 
\begin{cases} C\cdot\log\Bigl(\frac{1}{C(1-t)} \Bigr)& \text{for $T> 1-1/C$}\\
                         0                                         & \text{for $T\leq 1-1/C$}
                         \end{cases}
 \end{equation}

Now let ${\rm FP}(k)$ be the actual number of false positives up to and including step $k$.
They prove that for any  accumulation function $h$ bounded by a constant $C$,
\begin{equation}
{\rm E} \Bigl[ \frac{{\rm FP}(\hat k_h)}{C/\alpha+\hat k_h} \Bigr] \leq \alpha
\end{equation}
They demonstrate numerically that the HingeExp rule offers superior empirical performance compared to ForwardStop  (and other competitors) in many situations.

Importantly, this class of accumulation tests require that  the p-values $p_i$  are  $U(0,1)$ if the null hypothesis $H_i$ is true, and are independent of the non-null p-values.
This implies that the class of sequential pvalues proposed in this paper can be used as inputs to these tests.

@@

\section{Inference for One Step}
\label{sec:singleStep}

@@

\section{Computation}
\label{sec:computation}

@@

 Let $F_{\mu,\sigma^2}^{a,b}(t)$ be the  cumulative distribution function of the truncated normal.
 Finally let $T=|T_{j^*}|$.
Then by assuming independence, a  simple approximation to the max-t value is given by
\begin{equation}
1-\prod_{j\notin E} \{F_{0,||\eta||^2}^{v_j^-, v_j^+}(T)-   F_{0,||\eta||^2}^{v_j^-, v_j^+}(-T) \}
\label{eqn:approxpv}
\end{equation}
This approximation is exact if the predictors are uncorrelated.

Figure \ref{fig:approxpv} shows an example with $n=50$ and $p=10$ features having pairwise correlation 0.5.
There are two moderate signals.
Shown are the p-values from the ``exact'' numerical computation (with $100,000$ bootstrap samples)
and the approximation based on (\ref{eqn:approxpv}).
\begin{figure}[htp]
\centering
  \includegraphics[width=3in]{figs/approxpv.pdf}
  \caption{Exact max-t pvalues and approximation based on (\ref{eqn:approxpv}), for a simulated example
  with $n=50, p=10$.}
  \label{fig:approxpv}
\end{figure}
The approximation seems quite good. By Sidak's lemma, we might conjecture that the approximation is conservative, but we have no proof of this.

\section{Simulation: Sparse Linear Regression}
\label{sec:sparseReg}

@@

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Fri May  1 09:29:28 2015
\begin{table}[ht]
  \centering
  \begin{tabular}{llcccc}
    \hline
    Method & Stopping Rule & $p_{\text{screen}}$ & $\text{FWER}_{\text{mod}}$ an
    & $\text{FDR}_{\text{mod}}$ 
    & $\text{FDR}_{\text{var}}$ \\ 
    \hline
    Selected & Basic & .290 & .000 & .002 & .039 \\ 
    Selected & Forward & .559 & .027 & .020 & .066 \\ 
    Selected & Strong & .041 & .014 & .008 & .042 \\ 
    \hline
    Saturated & Basic & .000 & .000 & .000 & .028 \\ 
    Saturated & Forward & .014 & .000 & .000 & .030 \\ 
    Saturated & Strong & .000 & .000 & .000 & .032 \\ 
    \hline
    Knockoffs & & .000 & --- & --- & .231 \\ 
    \hline
  \end{tabular}
  \caption{\WFcomment{Comment on results once they are correct. These numbers are wrong right now.}}
\end{table}

\section{Principal Components Analysis}
\label{sec:pca}

\WFcomment{Can we do this one?}
Here we consider model selection for   principal components analysis. In that case we are given a data matrix $X \in \R^{n\times d}$, with which we form a sample %covariance matrix
\[
S = \frac{1}{n-1} \sum_{i=1}^n(x_i - \bar x)^2
\]
The first $d$ principal component loadings are the first $d$ eigenvectors of $S$, which call $u_1,\ldots, u_d$. These induce a sequence of nested Wishart models:
\[
M_0 \sub M_1 \sub \cdots \sub M_d
\]
in which
\begin{equation}
  M_k:\; (n-1) S \sim W_d\left(\lambda_0 I_d + \sum_{i=1}^k     \lambda_i u_i u_i', \;\;\; n-1\right).
\end{equation}
 By successively conditioning on $u_j, j<k$ one can obtain an exact sequential test
for the complete null hypothesis $H_0: \lambda_j=0 \; \forall j \neq k$.  The computation of the resulting p-values requires numerical integration.
The details are given iin \citet{choi2014selecting}.

\RTcomment{Jon- add more details?}

\section{Discussion}
\label{sec:pca}
